--- 
title: "Project Title"
author:
  - name: First Last
    affiliation: Department of<br>Vanderbilt University School of Medicine<br>MSCI Biostatistics II
date: last-modified
format:
  html:
    self-contained: true
    number-sections: true
    number-depth: 3
    anchor-sections: true
    smooth-scroll: true
    theme: journal
    toc: true
    toc-depth: 3
    toc-title: Contents
    toc-location: left
    code-link: false
    code-tools: true
    code-fold: show
    code-block-bg: "#f1f3f5"
    code-block-border-left: "#31BAE9"
    highlight-style: tango
    reference-location: margin
    fig-cap-location: margin

execute:
  warning: false
  message: false
---

```{r setup}
require(rms)    # also loads Hmisc package
getRs('abd.r')  # if using getabd function
```
# Background

To load this template into a running `RStudio` script editor window (usually the upper left pane), type the following in the `RStudio` command console (usually the lower left pane).

```{r eval=FALSE}
getRs('b2.qmd', put='rstudio')
```

# Data Import

If reading a `csv` dataset from ABD use the `getabd` function.  If reading a binary R file from `hbiostat.org/data` use the `Hmisc` `getHdata` function.[See [this](https://hbiostat.org/rflow/fcreate.html) for methods for importing your own datasets.]{.aside}

```{r import}
# getabd()     # run without an argument if you need a list of all
               # available ABD datasets
d <- getabd('02-e-2a') # fetch ?? csv file and store contents in data frame d
# getHdata(support) etc. for larger fully annotated datasets
# d <- support  # copy data frame produced by getHdata to use a short name
html(contents(d))  # data dictionary for imported dataset
```

# Descriptive Statistics

```{r desc}
html(describe(d))
```

# Computing Environment

The `session` function printed the following.  `session` is defined inside an object `markupSpecs` provided by `Hmisc`.

`r markupSpecs$html$session()`

