---
title: "Linear Model Analysis of Rosner `lead` Dataset"
author: "FE Harrell"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: pygments
    toc: yes
major: regression modeling
minor: ordinary least squares

---
# Data Description
```{r contents,results='asis'}
# results='asis' in the chunk header because html(contents) output html directly
knitrSet(lang='markdown')
require(rms)
getHdata(lead)
h <- html(contents(lead), file='', maxlevels=10, levelType='table')
```
```{r describe}
describe(lead)
```

# ANOVA for Three Exposure Groups
```{r anova}
dd <- datadist(lead); options(datadist='dd')
f <- ols(maxfwt ~ group, data=lead)
f
anova(f)
```

# ANCOVA for Three Groups Adjusted for Age
```{r ancova}
fa <- ols(maxfwt ~ age + group, data=lead)
fa
anova(fa)
# Also get group effect by subtracting SSRs or SSEs using anova(f)

# Continuous Lead Exposure Analyses
```{r contexp}
f <- ols(maxfwt ~ age + ld72 + ld73, data=lead)
f         # same as print(f)
coef(f)   # retrieve coefficients
g <- Function(f)  # create an R function that represents the fitted model
# Note that the default values for g's arguments are medians
g
# Estimate mean maxfwt at age 10, .1 quantiles of ld72, ld73 and .9 quantile of ld73
# keeping ld72 at .1 quantile
g(age=10, ld72=21, ld73=c(21, 47))  # more exposure in 1973 decreased y by 6
# Get the same estimates another way but also get std. errors
predict(f, data.frame(age=10, ld72=21, ld73=c(21, 47)), se.fit=TRUE)
```

# Diagnostics Based on Residuals
```{r resid}
r <- resid(f)
par(mfrow=c(2,2))   # 2x2 matrix of plots
plot(fitted(f), r); abline(h=0)  # yhat vs. r
with(lead, plot(age,  r));    abline(h=0)
with(lead, plot(ld73, r));    abline(h=0)
qqnorm(r)           # linearity indicates normality
```

# Partial Effects of Predictors
```{r peffects}
ggplot(Predict(f))
ggplot(Predict(f, age))   # plot only age effect, using default range,
                          # 10th smallest to 10th largest age
ggplot(Predict(f, age=3:15))  # plot age=3,4,...,15
ggplot(Predict(f, age=seq(3,16,length=150)))   # plot age=3-16, 150 points
# Get confidence limits for individual response
ggplot(Predict(f, age, conf.type='individual'))
# Show both types of 0.99 confidence limits on one plot
p1 <- Predict(f, age, conf.int=0.99, conf.type='individual')
p2 <- Predict(f, age, conf.int=0.99, conf.type='mean')
p <- rbind(Individual=p1, Mean=p2)
ggplot(p)

# Non-plotted variables are set to reference values (median and
#  mode by default). To control the settings of non-plotted values use e.g.
ggplot(Predict(f, ld73, age=3))

# To make separate lines for two ages:
ggplot(Predict(f, ld73, age=c(3,9)))  # add ,conf.int=FALSE to suppress conf. bands
# To plot a 3-d surface for two continuous predictors against
#  yhat; color coding for predicted mean maxfwt
bplot(Predict(f, ld72, ld73))
```

# Point Estimates for Partial Effects
The `summary` function can compute point estimates and confidence
intervals for effects of individual predictors, holding other
predictors to selected constants.  The constants you hold other
predictors to will only matter when the other predictors interact with
the predictor whose effects are being displayed.

How predictors are changed depend on the type of predictor:
- Categorical predictors: differences against the reference (most
  frequent) cell by default
- Continuous predictors: inter-quartile-range effects by default

The estimated effects depend on the type of model:
- `ols`: differences in means
- logistic models: odds ratios and their logs
- Cox models: hazard ratios and their logs
- quantile regression: differences in quantiles

```{r summary}
summary(f)         # inter-quartile-range effects
summary(f, age=5)  # adjust age to 5 when examining ld72,ld73
                   # (no effect since no interactions in model)
summary(f, ld73=c(20, 40))  # effect of changing ld73 from 20 to 40
```
There is a `plot` method for `summary` results.  By default it
shows 0.9, 0.95, and 0.99 confidence limits.
```{r plotsummary,top=2}
# top=2 in chunk header: leave extra room for labels
plot(summary(f))
```

# Getting Predicted Values
## Using `predict`
```{r predict}
predict(f, data.frame(age=3, ld72=21, ld73=21))
# must specify all variables in the model

predict(f, data.frame(age=c(3, 10), ld72=21, ld73=c(21, 47)))
# predictions for (3,21,21) and (10,21,47)

newdat <- expand.grid(age=c(4, 8), ld72=c(21, 47), ld73=c(21, 47))
newdat
predict(f, newdat)     # 8 predictions

predict(f, newdat, conf.int=0.95)  # also get CLs for mean
predict(f, newdat, conf.int=0.95, conf.type='individual')  # CLs for indiv.
```
## The brute-force way
```{r predmanual}
# Model is b1 + b2*age + b3*ld72 + b4*ld73
b <- coef(f)
# For 3 year old with both lead exposures 21
b[1] + b[2]*3 + b[3]*21 + b[4]*21
```

## Using `Function` function
```{r Function}
g <- Function(f)
g(age=c(3, 8), ld72=21, ld73=21)       # 2 predictions
g(age=3)              # 3 year old at median ld72, ld73
```

# General ANOVA
- Use `anova(fitobject)` to get all total effects and
  individual partial effects
- Use `anova(f,age,sex)` to get combined partial effects of
  `age` and `sex`, for example
- Store result of `anova` in an object in you want to print it
  various ways, or to plot it:
  
```{r printanova}
an <- anova(f)
an                     # same as print(an)
print(an, 'names')     # print names of variables being tested
print(an, 'subscripts')# print subscripts in coef(f) (ignoring
                       # the intercept) being tested
print(an, 'dots')      # a dot in each position being tested
anova(f, ld72, ld73)   # combine effects into a 2 d.f. test
```

# Computing Environment
```{r rsession,echo=FALSE}
si <- sessionInfo(); si$loadedOnly <- NULL
print(si, locale=FALSE)
```
```{r cite,results='asis',echo=FALSE}
print(citation(), style='text')
```
