--- 
title: "NHANES"
author:
  - name: Frank Harrell
    affiliation: Department of Biostatistics<br>Vanderbilt University School of Medicine<br>MSCI Biostatistics II
date: last-modified
format:
  html:
    self-contained: true
    number-sections: true
    number-depth: 3
    anchor-sections: true
    smooth-scroll: true
    theme: journal
    toc: true
    toc-depth: 3
    toc-title: Contents
    toc-location: left
    code-link: false
    code-tools: true
    code-fold: show
    code-block-bg: "#f1f3f5"
    code-block-border-left: "#31BAE9"
    reference-location: margin
    fig-cap-location: margin

execute:
  warning: false
  message: false
---

```{r setup}
require(rms)
options(prType='html')
```

# Problem: How Are Demographics Associated with Height?

* Dataset: `nhgh`
* N observations: 6795
* Response: height
* Predictors: Race/ethnicity, sex, age

# Data Import

```{r import,results='asis'}
getHdata(nhgh)
d <- upData(nhgh, keep=.q(sex, age, re, ht, SCr, gh))
html(contents(d))
```

# Descriptive Statistics

```{r desc, results='asis'}
html(describe(d))
```

# Plot Raw Data

There are many tied values of `ht` so use hexagonal binning to depict frequency counts of coincident points.  Use `ggplot2` faceting to stratify by `re` and `sex`.  Make the plot taller than usual with chunk markup `#| fig.height: 8.5` (8.5 inches).

```{r}
#| fig.height: 8.5
hlab <- function(x) {
  x <- as.character(substitute(x))
  label(d[[x]], plot=TRUE, default=x)
}
cuts <- c(1:5, 10, 15, 20, 30)  # cutpoints for cell N
ggplot(d, aes(x=age, y=ht)) +
  facet_grid(re ~ sex) +
  stat_binhex(aes(alpha=..count.., color=cut2(..count.., cuts=cuts)), bins=80) +
  xlab(hlab(age)) + ylab(hlab(ht)) +
  guides(alpha='none', fill=FALSE, color=guide_legend(title='Frequency'))
```

# Linear Model

* No assumption of linearity for continuous predictors
* Allow for age $\times$ sex interaction (different shapes of age relationship by sex)

```{r results='asis'}
dd <- datadist(d); options(datadist='dd')
# x=TRUE needed for studentized residuals
f <- ols(ht ~ re + sex * rcs(age, 4), data=d, x=TRUE)
f
latex(f)
```

# Diagnostics

## Residuals

```{r}
r <- resid(f, type='studentized')
ggplot(mapping=aes(fitted(f), r)) + geom_point(alpha=.2)
ggplot(mapping=aes(sample=r)) + geom_qq() + stat_qq_line() +
  xlab('Studentized Residuals') + ylab('Normal Quantiles')
```

* What is your opinion about which model assumptions hold and which don't?
* How might the right hand side of the model not fit adequately?

## Influence Statistics

Find observations that if removed would change a parameter by more than 0.15 standardized (by standard error) units.[Because of the large sample size there were no overly influential observations at standard cutoffs such as 0.4.  The cutoff was relaxed to 0.15 for illustration.]{.aside}  `which.influence` in `rms` recognizes that some predictors have multiple $\beta$s associated with them (e.g., splined predictors or categorical predictors like `re` with > 2 levels).  It finds observations that are influential on *any* of the predictor's parameters.

**NOTE**: The output of `show.influence` will be incorrect for you until `rms` 6.4-2 is released.  The output below is correct.

```{r}
w <- which.influence(f, 0.15)
w
show.influence(w, d)
```

# Computing Environment

`r markupSpecs$html$session()`
