---
title: "RMS Chapter 2 Example: Analysis of FEV Dataset"
author: "Frank Harrell"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: pygments
    toc: yes
major: regression modeling; rms
minor: curve fitting
---
```{r setup}
require(Hmisc)
knitrSet(lang='markdown')
```
# Descriptive Statistics
```{r importdesc}
require(rms)
getHdata(FEV)
contents(FEV)
describe(FEV)
dd <- datadist(FEV); options(datadist='dd')
dd
```

# Regression Models of Increasing Complexity
FEV is log-transformed to satisfy normality and equal variance assumptions.

## Single Predictor, Linear Effect
```{r lin,w=6.5,mfrow=c(1,2)}
f <- ols(log(fev) ~ age, data=FEV)
f
r <- as.numeric(resid(f))
with(FEV, plot(age, r))
qqnorm(r)
qqline(r)
```
```{r linplot}

# Show algebraic form of fitted model
Function(f)
g <- Function(f)
exp(g()); exp(g(age=10))   # Evaluate the fitted model, antilog to get original scale
ggplot(Predict(f)) + geom_point(aes(x=age, y=log(fev), color=sex), FEV)
```

## Restricted Cubic Spline With 3 Default Knots
The knots are at these quantiles of age: 0.1, 0.5, 0.9.
```{r rcs3}
f <- ols(log(fev) ~ rcs(age, 3), data=FEV, x=TRUE)
f
g <- Function(f)
g(10:20)
g()
g
ggplot(Predict(f))
anova(f)
```

## RCS With 5 Default Knots
```{r rcs5}
f <- ols(log(fev) ~ rcs(age, 5), data=FEV)
f
Function(f)
anova(f)
ggplot(Predict(f))
```

## RCS with 5 Knots, Additive (Non-Interacting) Sex Effect
```{r rcs5sex,w=5}
f <- ols(log(fev) ~ rcs(age, 5) + sex, data=FEV)
Function(f)
anova(f)
ggplot(Predict(f, age, sex)) + geom_point(aes(x=age, y=log(fev), color=sex), FEV)
```

## RCS in Age Interacting With Sex
The following model allows for different shapes of effects for males and females.
```{r rcsia}
f <- ols(log(fev) ~ rcs(age, 5) * sex, data=FEV)
Function(f)
anova(f)
ggplot(Predict(f, age, sex))
```

## RCS with Age * Sex and Additive Nonlinear Effect of Height
We show the joint age and height effects using a color image plot
```{r rcsiah}
f <- ols(log(fev) ~ rcs(age, 5) * sex + rcs(height, 5), data=FEV)
anova(f)
bplot(Predict(f, age, height))
```

## Model Ignoring Sex and Height But Including Smoking History
By plotting spike histograms showing the smoking-specific age distribution we see that there are almost no very young children who have smoked.  This limits the power to detect an age by smoking interaction.
```{r amoke,w=5}
f <- ols(log(fev) ~ rcs(age, 5) + smoke, data=FEV)
ggplot(Predict(f, age, smoke), rdata=FEV)
```

# Computing Environment
```{r rsession,echo=FALSE}
si <- sessionInfo(); si$loadedOnly <- NULL
print(si, locale=FALSE)
```
```{r cite,results='asis',echo=FALSE}
print(citation(), style='text')
```
